---
title: "project-503"
author: "Mohammad"
date: "2023-06-03"
output:
 
  html_document: default
---
```{r}
knitr::opts_chunk$set(echo = TRUE)

library(mlbench)
library(Hmisc)
library(e1071)
library(caret)
##library(tidyr)
library(corrplot)
library(AppliedPredictiveModeling)
library(car)
library(lattice)
library(lars)
library(stats)
library(pls)
library(dplyr)
library(kernlab)
library(randomForest)
library(gbm)
library(earth)
library(plotmo)
library(plotrix)
library(TeachingDemos)
```



```{r}
#load heart disease dataset
heart_df <- read.table(file.choose(), header=TRUE, sep=",")
head(heart_df)

#lets see how dataset look like
str(heart_df)

summary(heart_df)

```


```{r}
#checking the NA values
sum(is.na(heart_df))


```


```{r}
# check Distinct values


heart_df %>% 
  summarise(n_age = n_distinct(age), n_sex = n_distinct(sex),
            n_chestpain = n_distinct(chest.pain.type),
            n_restbp=n_distinct(resting.blood.pressure),
            n_chol = n_distinct(serum.cholestoral), 
            n_fastbs = n_distinct(fasting.blood.sugar),
            n_restecg = n_distinct(resting.electrocardiographic.results), 
            n_HR= n_distinct(max.heart.rate),
            n_exercise = n_distinct(exercise.induced.angina), 
            n_oldpeak = n_distinct(oldpeak),
            n_STsegment = n_distinct(ST.segment), 
            n_mvessels = n_distinct(major.vessels), 
            n_thal = n_distinct(thal),
            n_heartdisease = n_distinct(heart.disease))




```

```{r}
#age distribution vs heart disease plot
heart_df %>% group_by(age, heart.disease) %>% summarise(count = n()) %>%
  ggplot() + geom_bar(aes(age, count,   fill = as.factor(heart.disease)), stat = "Identity") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, size = 10)) + 
  ylab("Count") + xlab("Age") + labs(fill = "heart.disease")
```
Blue plot which is the presence of heart disease is left skewed which shows age has a positive correlation with heart disease. 

```{r}
#boxplot to displays the age distribution of heart diagnosis
boxplot(heart_df$age ~ heart_df$heart.disease,
        main="Heart disease distribution by Age",
         ylab="Age",xlab="Heart disease")
```

```{r}
#Gender analysis
#The proportion of females and males patients in the dataset.

heart_df %>% 
    group_by( sex ) %>% 
    summarise( percent = 100 * n() / nrow( heart_df ))
```


There are 32.2 % females and 67.8% males in the dataset



```{r}
#Check the percentage of males and females with heart disease

female_yes <- table(heart_df[heart_df$sex==0,]$heart.disease)
male_yes <- table(heart_df[heart_df$sex==1,]$heart.disease)
FMcombine_yes <- rbind(female_yes,male_yes)

#Rename columns names and rows names.
colnames(FMcombine_yes) <- c("Yes.disease", "No.disease")
rownames(FMcombine_yes) <- c("Females", "Males")

#Display the table
FMcombine_yes
```
There are 67 females out of 87 who have diagnosed with heart disease and 83 males out of 183 were diagnosed with heart disease.

```{r}
#mosaic plot for gender vs heart.disease
mosaicplot(heart_df$sex ~ heart_df$heart.disease,
           main="Heart disease vs Gender", shade=FALSE,color=blues9,
           xlab="Gender", ylab="Heart disease")

```

```{r}
#chest pain type analysis
#heart_df$chest.pain.type <- as.factor(heart_df$chest.pain.type)

ggplot(data = heart_df, aes(x = heart.disease, fill = as.factor(chest.pain.type))) + 
  geom_bar(position = "fill") +
  labs(title = "Heart disease Distributions by Chest pain type",
       x = "Heart disease",
       y = "chest pain type") +
  theme_test()
```


```{r}

#Rest Blood Pressure distribution vs heart disease plot
boxplot(heart_df$resting.blood.pressure ~ heart_df$heart.disease,
        main="Heart disease distribution by resting.blood.pressure",
         ylab="resting.blood.pressure",xlab="Heart disease")
```

```{r}
#major blood vessels vs heart.disease analysis
#heart_df$major.vessels <- as.factor(heart_df$major.vessels)

ggplot(data = heart_df, aes(x = heart.disease, fill = as.factor(major.vessels))) + 
  geom_bar(position = "fill") +
  labs(title = "Heart disease Distributions by Number of major vessels ",
       x = "Heart disease",
       y = "Number of major vessels") +
  theme_test()
```

```{r}
#serum.cholestoral distribution vs heart disease plot

boxplot(heart_df$serum.cholestoral ~ heart_df$heart.disease,
        main="Heart disease distribution by serum.cholestoral",
         ylab="serum.cholestoral",xlab="Heart disease")
```


```{r}
#max.heart.rate distribution vs heart disease plot

boxplot(heart_df$max.heart.rate ~ heart_df$heart.disease,
        main="Heart disease distribution by max.heart.rate",
         ylab="max.heart.rate",xlab="Heart disease")

```

```{r}

```






```{r}
#correlations
corr <- cor(heart_df[,1:13])

round(corr,2)
```

```{r}
#plot correlations
corrplot::corrplot(cor(heart_df[, 1:13]))
```

```{r}
#split dataset
set.seed(502)
trainingrows <- createDataPartition(heart_df$heart.disease, p=0.8, list=FALSE)
heart_train <- heart_df[trainingrows,]
heart_test <- heart_df[-trainingrows,]

#preprocess including center and scale
heart_trainimp <- preProcess(heart_train, "knnImpute")
heart_trainpredict <- predict(heart_trainimp, heart_train)
heart_testpredict <- predict(heart_trainimp, heart_test)


summary(heart_trainpredict)
```
##Linear Discriminant Analysis

```{r}

#LDA
lda_fit <- train(as.factor(heart.disease) ~ ., method = "lda", data = heart_train)
lda_fit
lda_predict <- predict(lda_fit, heart_test)


confusionMatrix(lda_predict, as.factor(heart_test$heart.disease))
```
##Logistic Regression

```{r}

set.seed(503)
lr_fit <- train(as.factor(heart.disease) ~ ., method = "glm", data = heart_train)
lr_fit
lr_predict <- predict(lr_fit, heart_test)
confusionMatrix(lr_predict, as.factor(heart_test$heart.disease))
```

```{r}
#rpart 
set.seed(503)
rpart_fit <- train(as.factor(heart.disease) ~ ., method = "rpart", data = heart_train)
rpart_fit
rpart_predict <- predict(rpart_fit, heart_test)
confusionMatrix(rpart_predict, as.factor(heart_test$heart.disease))
```

```{r}
library(rpart)

library(rpart.plot)
rpart.plot(rpart_fit$finalModel,   
           type=5,
           fallen.leaves = FALSE,
           box.palette = "GnRd",
           nn=TRUE)
```



##KNN
```{r}
ctrl <- trainControl(method = "cv", verboseIter = FALSE, number = 5)
set.seed(503)
knn_fit <- train(as.factor(heart.disease) ~ ., 
                data = heart_train, method = "knn", preProcess = c("center","scale"),
                trControl = ctrl , tuneGrid = expand.grid(k = seq(1, 20, 2)))

knn_fit
knn_predict <- predict(knn_fit, heart_test)
confusionMatrix(knn_predict, as.factor(heart_test$heart.disease))
```




##Support Vector Machine (SVMLinear)
```{r}
ctrl <- trainControl(method = "cv", verboseIter = FALSE, number = 5)
set.seed(503)

grid_svm <- expand.grid(C = c(0.01, 0.1, 1, 10, 20))


svm_fit <- train(as.factor(heart.disease) ~ .,data = heart_train,
                 method = "svmLinear", preProcess = c("center","scale"),
                 tuneGrid = grid_svm, trControl = ctrl)
svm_fit
svm_predict <- predict(svm_fit, heart_test)
confusionMatrix(svm_predict, as.factor(heart_test$heart.disease))
```
### Support Vector Machines with Radial kernel
```{r}

sigmaEst <- kernlab::sigest(as.matrix(heart_train[,1:13]))
svmgrid <- expand.grid(sigma = sigmaEst, C = 2^seq(-4,+4))

set.seed(503)
svmR_fit <- train(as.factor(heart.disease) ~ .,data = heart_train,
                 method = "svmRadial", preProcess = c("center","scale"),
                 tuneGrid = svmgrid, trControl = ctrl)
svmR_fit
svmR_predict <- predict(svmR_fit, heart_test)
confusionMatrix(svmR_predict, as.factor(heart_test$heart.disease))
```


##Random Forest
```{r}
control<- trainControl(method = "cv", number = 5, verboseIter = FALSE)
grid <-data.frame(mtry = seq(1, 10, 2))
set.seed(503)
rf_fit <- train(as.factor(heart.disease) ~ ., method = "rf", data = heart_train, ntree = 20, trControl = control, tuneGrid = grid)
rf_fit

rf_predict <- predict(rf_fit, heart_test)
confusionMatrix(rf_predict, as.factor(heart_test$heart.disease))
```


##QDA
```{r}
set.seed(503)
qda_fit <- train(as.factor(heart.disease) ~ ., method = "qda", data = heart_train)
qda_fit

qda_predict <- predict(qda_fit, heart_test)
confusionMatrix(qda_predict, as.factor(heart_test$heart.disease))
```



##Gradient Boosting Machine
```{r}
gbmGrid <-  expand.grid(interaction.depth = c(1, 5, 10, 25, 30),
                        n.trees = c(5, 10, 25, 50),
                        shrinkage = c(0.1, 0.2, 0.3,  0.4, 0.5),
                        n.minobsinnode = 20)

set.seed(503)
gbm_fit <- train(as.factor(heart.disease) ~ ., method = "gbm", data = heart_train,  trControl = control, verbose = FALSE, tuneGrid = gbmGrid)
gbm_fit
gbm_predict <- predict(gbm_fit, heart_test)
confusionMatrix(gbm_predict, as.factor(heart_test$heart.disease))
```

##Bagged trees
```{r}
set.seed(503)
bagged_fit <- train(as.factor(heart.disease) ~ ., method = "treebag", nbagg=50, 
                    data = heart_train,  trControl = control, metric="Accuracy")

bagged_fit
bagged_predict <- predict(bagged_fit, heart_test)
confusionMatrix(bagged_predict, as.factor(heart_test$heart.disease))
```
##Neural network
```{r}
set.seed(503)

nnetGrid <- expand.grid(size=1:3, decay=c(0,0.1,0.2,0.3,0.4,0.5,1,2))

nnet_fit <- train(as.factor(heart.disease) ~ ., method = "nnet", 
                    data = heart_train, tuneGrid=nnetGrid,
                  trace=FALSE, maxit=2000, trControl = control, metric="Accuracy")

nnet_fit

nnet_predict <- predict(nnet_fit, heart_test)
confusionMatrix(nnet_predict, as.factor(heart_test$heart.disease))
```




Support vector machine with svmlinear kernel has the best performance among all.





```{r}

nscGrid <- data.frame(threshold = seq(0,25, length=30)) 
set.seed(503) 

nsc_fit <- train(as.factor(heart.disease) ~ .,method = "pam",data = heart_train,
                 preProc = c("center", "scale"), tuneGrid = nscGrid,trControl = ctrl,metric = "Accuracy")

nsc_fit

nsc_predict <- predict(nsc_fit, heart_test)
confusionMatrix(nsc_predict, as.factor(heart_test$heart.disease))
```

```{r}
set.seed(503)
mda_fit <- train(x = heart_train[, 1:13],
                 y = as.factor(heart_train$heart.disease),
                 method = "mda",
                 tuneGrid = expand.grid(subclasses = 1:3),
                 metric = "Accuracy",
                 trControl = ctrl)
mda_fit

mda_predict <- predict(mda_fit, heart_test)
confusionMatrix(mda_predict, as.factor(heart_test$heart.disease))

```





```{r}
#top important variables of SVM
varImp(svm_fit, top=5)
plot(varImp(svm_fit, top=5))
```



```{r}
plot(svmR_fit, scales = list(x = list(log = 2))) 
```









